---
title: "R Notebook reproduce figures from BioDataManifest manuscript"
output: html_notebook
---

This notebook accompanies Bottomly et al. "Biomedical Data Manifest: A lightweight data documentation mapping to increase transparency for AI/ML".  Submitted to Scientific Data.

Assumes the supplemental table Excel file (`SuppTables-06-20-2025.xlsx`) is in the current directory.

```{r message=FALSE}

library(tidyverse)
library(ggplot2)
library(patchwork)
library(readxl)
library(ggrepel)

survey_res <- read_excel("SuppTables-06-20-2025.xlsx", sheet="Table S5")

```

# Initially parse survey

```{r}

cur_question <- names(survey_res)[1]

question_groups <- tribble(
    ~question, ~column_idx, ~statement,
    cur_question, 1, as.character(unname(unlist(survey_res[1,1])))
)

question_order <- c(cur_question)

for(i in 2:ncol(survey_res)){
    
    if (grepl("^\\.\\.\\.\\d+$", names(survey_res)[i]) == F){
        cur_question <- names(survey_res)[i]
        question_order <- c(question_order, cur_question)
    }
    
    question_groups <- bind_rows(
        question_groups,
        tribble(
        ~question, ~column_idx, ~statement,
        cur_question, i, as.character(unname(unlist(survey_res[1,i])))
    ))
}

question_groups <- inner_join(
    question_groups,
    as_tibble(question_order) %>%
        rename(question = "value") %>%
        mutate(question_idx=seq_along(question)-1),
    by="question"
)

question_groups <- filter(question_groups, column_idx > 2)

question_groups <- group_by(question_groups, question)

reformatted_questions <- lapply(group_split(question_groups), function(group_tbl){
    
    tmp_wide_question <- select(survey_res, all_of(c(1, 2, group_tbl$column_idx)))
    tmp_wide_question[1,1] <- "respondent"
    tmp_wide_question[1,2] <- "primary_role"
    names(tmp_wide_question) <- tmp_wide_question[1,] %>% 
        unlist %>% 
        unname
    
    if (ncol(tmp_wide_question) >= 4 && all(names(tmp_wide_question)[seq(4, ncol(tmp_wide_question), by=2)] == "Other (please specify)")){
        names(tmp_wide_question)[seq(4, ncol(tmp_wide_question), by=2)] <- paste(names(tmp_wide_question)[seq(3, ncol(tmp_wide_question), by=2)], "comment", sep="_")
        names(tmp_wide_question)[seq(3, ncol(tmp_wide_question), by=2)] <- paste(names(tmp_wide_question)[seq(3, ncol(tmp_wide_question), by=2)], "response", sep="_")
    }else{
        tmp_wide_question[[paste(names(tmp_wide_question)[3], "comment", sep="_")]] <- NA_character_
        names(tmp_wide_question)[3] <- paste(names(tmp_wide_question)[3], "response", sep="_")
    }
    
    pivot_longer(data=tmp_wide_question[-1,], 
                 cols=-c(1:2),
                 names_sep= "_",
                 names_to=c("statement", "type"), 
                 values_to="response") %>%
        mutate(question=group_tbl$question[1],
               question_idx=group_tbl$question_idx[1]) %>%
        pivot_wider(names_from=type, values_from=response) %>%
        left_join(select(group_tbl, statement, column_idx), by="statement")
    
}) %>%
    bind_rows

reformatted_questions <- mutate(reformatted_questions, 
                                role_category=ifelse(primary_role %in% c("Data Manager", "Computationalist"),
                                                     "DM/Comp", "Bench/Clinical"))

#address the issue of choosing Other for questions 2 and 3
##change statement to Response and add 'Other:' to the response value

reformatted_questions <- reformatted_questions %>% 
    mutate(response = ifelse(statement == "Response" & response == "Other (please specify)", paste("Other:", comment), response))

```

# Add in annotation about 'gateway questions'

```{r}

reformatted_questions <- mutate(reformatted_questions, can_answer="yes")

#"Have not heard about any of these concepts before"--skips rest of survey

no_concepts <- reformatted_questions %>% 
    filter(question_idx == 2 & 
               response == "Have not heard about any of these concepts before")

reformatted_questions <- mutate(reformatted_questions, 
                                can_answer = ifelse(respondent %in% no_concepts$respondent & question_idx > 2, 
                                                    "no", 
                                                    can_answer))

#"Do you conduct human subjects or non-human subjects research?"
#"Non-human subjects only"--skips ELSI
non_human <- reformatted_questions %>% 
    filter(question_idx == 7 & response == "Non-human subjects only")

reformatted_questions <- mutate(reformatted_questions, 
                                can_answer = ifelse(respondent %in% non_human$respondent & question_idx == 8, 
                                                    "no", 
                                                    can_answer))

#"As part of your research do you utilize expert labelers to annotate the dataset (e.g. radiologists, pathologists, domain experts annotating images, or text for NLP)."
#"No"--skips Labeling Provenance and Lineage

no_labels <- reformatted_questions %>% 
    filter(question_idx == 10 & response == "No")

reformatted_questions <- mutate(reformatted_questions, 
                                can_answer = ifelse(respondent %in% no_labels$respondent & question_idx == 11, 
                                                    "no", 
                                                    can_answer))

#"After you disseminate/release a dataset do you actively update it?"
#"No"--ends survey

no_update <- reformatted_questions %>% 
    filter(question_idx == 12 & response == "No")

reformatted_questions <- mutate(reformatted_questions, 
                                can_answer = ifelse(respondent %in% no_update$respondent & question_idx == 13, 
                                                    "no", 
                                                    can_answer))

```

# Figure 2

```{r fig.width = 10, fig.height = 5}

uniq_roles <- filter(reformatted_questions, question_idx == 2) %>%
    select( respondent, primary_role, role_category, response) %>%
    unique()

p1 <- ggplot(data=uniq_roles, mapping=aes(x=primary_role, fill=response)) +
    geom_bar() +
    scale_fill_brewer(type="qual",palette="Set1", direction=-1, name = "Heard of common data management/curation concepts?") +
    facet_wrap(~role_category, scales="free_x") +
    theme_bw() + xlab("")

p1

```

# Figure 3

```{r}

data_handling <- filter(reformatted_questions, question_idx %in% 3:4 & can_answer == "yes")

data_handling <- mutate(data_handling,
                        resp_fac=factor(response, levels=c("Myself", 
                                                           "Someone else (a dedicated person in the lab/team)", 
                                                           "Split across multiple people",
                                                           "Rely on software logs /tools to capture this",
                                                           "Other: Does not apply"), 
                                        labels=c("Myself", 
                                                 "Dedicated person", 
                                                 "Multiple people",
                                                 "Software",
                                                 "Does not apply"),
                                        ordered = T))

question_tbl <- tribble(
    ~question, ~short_question,
    "Who is responsible for documenting how the data was generated in your lab/team after data generation is complete in order to disseminate or share the data?", "Who provides documentation while data is being generated?",
    "Who is responsible for documenting how the data was generated in your lab/team while the data is being generated?", "Who provides documentation for dissemination?"
)

data_handling <- inner_join(data_handling, question_tbl, by="question")

color_map <- setNames(RColorBrewer::brewer.pal(n=5, name="Dark2"), 
                      c("Myself", "Dedicated person", "Multiple people", "Software", "Does not apply"))

plot_list <- lapply(group_split(data_handling, .by=question_idx), function(x){
    ggplot(data=x, mapping=aes(x=resp_fac, fill=resp_fac)) +
        geom_bar() +
        facet_wrap(~role_category) +
        scale_x_discrete(drop=F) +
        scale_fill_manual(values=color_map, guide="none") +
        theme_bw() + 
        theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
        xlab("") +
        ggtitle(x$short_question[1])
})

wrap_plots(plot_list, ncol=1)

```

# Figure S2

```{r fig.width = 14, fig.height = 5}

response_scores <- tribble(
    ~response, ~score,
    "Essential", 5,
    "Less Essential", 3,
    "Non-Essential or Redundant", 0,
    "Unclear Meaning", NA)

reformatted_questions <- inner_join(reformatted_questions, response_scores, by="response")

#make a shorthand for questions

topics <- c("Ethical, Legal and Social Issues (ELSI)", "Dataset Composition",
            "Maintenance and Distribution", "Labeling Provenance and Lineage", 
            "Provenance and Lineage in datasets", "Uses of Data")

reformatted_questions <- mutate(reformatted_questions, topic=NA_character_)

for(cur_topic in topics){
    
    reformatted_questions <- mutate(reformatted_questions, topic=ifelse(grepl(cur_topic, question, fixed=T), cur_topic, topic))
}

statement_summary_total <- summarize(reformatted_questions, 
                                     OverallMean=mean(score, na.rm=T), 
                                     nclear=sum(is.na(score)==F), 
                                     n=n(), 
                                     .by=c("topic", "statement"))

statement_summary_total <- mutate(statement_summary_total,
                                     prop_clear = nclear / n)

p1 <- ggplot(data=statement_summary_total, mapping=aes(x=topic, y=OverallMean)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(height=0, width=.25, mapping=aes(color=prop_clear, size=n)) +
    geom_hline(yintercept=4, linetype = "dashed") +
    scale_colour_viridis_c(direction = -1) +
    theme_bw() + xlab("") + ylab("Avg. Score")


p1

```

# Figure 4

```{r}

statement_summary_category <- summarize(reformatted_questions, 
                                        Mean=mean(score, na.rm=T), 
                                        nclear=sum(is.na(score)==F), 
                                        n=n(),
                                        .by=c("topic", "statement", "role_category"))

statement_summary_category <- mutate(statement_summary_category,
                                     prop_clear = nclear / n)

p2 <- ggplot(data=statement_summary_category, mapping=aes(x=role_category, y=Mean)) +
    geom_boxplot() +
    geom_jitter(height=0, width=.25, mapping=aes(color=prop_clear, size=n)) +
    geom_hline(yintercept=4, linetype = "dashed") +
    scale_colour_viridis_c(direction = -1) +
    facet_wrap(~topic, scales="free_y") +
    theme_bw() + xlab("") + ylab("Score Summary")

p2

```

# Further summarize the survey (basis of Table S6) 

```{r}

statement_summary_category <- summarize(reformatted_questions, 
                                        AvgScore=mean(score, na.rm=T), 
                                        n_clear=sum(is.na(score)==F), 
                                        n=n(),
                                        .by=c("topic", "statement", "role_category")) %>%
    mutate(should_include = ifelse(AvgScore >= 4, "yes", "no"),
           PercClear = n_clear / n) %>%
    pivot_wider(id_cols=c("topic", "statement"), names_from="role_category", values_from=c("AvgScore", "PercClear", "should_include"))


comment_summary <- pivot_wider(reformatted_questions,
        id_cols=c("question_idx", "column_idx", "topic","statement"),
        names_from = respondent, values_from = comment)


survey_summary <- inner_join(statement_summary_category, comment_summary, by=c("topic", "statement"))

```

# Figure 5a/b

```{r}

survey_summary <- mutate(survey_summary,
                          score_cat = case_when(
                              `should_include_Bench/Clinical` == "yes" & `should_include_DM/Comp` == "yes" ~ "Both",
                              `should_include_Bench/Clinical` == "no" & `should_include_DM/Comp` == "no" ~ "Neither",
                              `should_include_Bench/Clinical` == "yes" & `should_include_DM/Comp` == "no" ~ "Only Bench/Clinical",
                              `should_include_Bench/Clinical` == "no" & `should_include_DM/Comp` == "yes" ~ "Only DM/Comp",
                          ))

#make a table of points to highlight indicated by letter

highlight_points <- filter(survey_summary,
                          (`AvgScore_Bench/Clinical` == 5.0 &  `AvgScore_DM/Comp` == 5.0) | 
                              (`AvgScore_Bench/Clinical` < 3.0 &  `AvgScore_DM/Comp` >= 4.0) |
                              (`AvgScore_Bench/Clinical` >= 4.0 &  `AvgScore_DM/Comp` < 3.75))

highlight_points <- arrange(highlight_points,
                            score_cat,
                            desc(mapply(function(x, y) x+y, `AvgScore_Bench/Clinical`, `AvgScore_DM/Comp`))) %>%
    mutate(letter=LETTERS[seq_along(statement)])

#convert the x-axis to a tibble and merge

label_pos <- tribble(
    ~score_cat, ~x, ~y,
    "Both", 5, 5.5,
    "Neither", 2, 1.5,
    "Only Bench/Clinical", 5, 1.5,
    "Only DM/Comp", 2, 5.5
)

use_cols <- setNames(scales::hue_pal()(4), label_pos$score_cat)

p1 <- ggplot(data=survey_summary, mapping=aes(x=`AvgScore_Bench/Clinical`, y=`AvgScore_DM/Comp`, fill=score_cat)) +
    geom_point(size=3, shape=21) +
    geom_vline(xintercept = 4.0, linetype="dashed") +
    geom_hline(yintercept = 4.0, linetype="dashed") +
    geom_text(data=label_pos, mapping=aes(x=x, y=y, label=score_cat, color=score_cat)) +
    scale_color_manual(values = use_cols,guide="none") +
    scale_fill_manual(values = use_cols,guide="none") +
    geom_label_repel(data=highlight_points, mapping=aes(label=letter, color=score_cat), min.segment.length = 0, box.padding = 0.5, fill="white") +
    ylim(c(1.25, 5.5)) +
    xlim(c(1, 5.5)) +
    theme_bw() +
    xlab("Avg. Score Bench/Clinical") +
    ylab("Avg. Score_DM/Comp")

#categorize the percent clear values as there are a lot of overlapping values to plot directly 
kept_fields <- survey_summary %>%
    mutate(`pclear_bin_Bench/Clinical` = cut(`PercClear_Bench/Clinical`*100, breaks=c(0, 70, 95, 100)),
           `pclear_bin_DM/Comp` = cut(`PercClear_DM/Comp`*100, breaks=c(0, 70, 95, 100))) 

kept_fields <- left_join(kept_fields, select(highlight_points, letter, statement), by="statement")

kept_fields <- mutate(kept_fields, letter = ifelse(is.na(letter), "", letter))

pos <- position_jitter(width=.4, height=.4, seed = 12232)

p2 <- ggplot(data=kept_fields, mapping=aes(x=`pclear_bin_Bench/Clinical`, y=`pclear_bin_DM/Comp`)) +
    geom_tile(fill="white", color="black", na.rm=F) +
    geom_label_repel(position=pos, mapping=aes(label=letter, color=score_cat), min.segment.length = 0, box.padding = 0.5, fill="white", max.overlaps = 10000) +
    geom_point(position=pos, shape=21, size=3, mapping=aes(fill=score_cat)) +
    annotate("text", x=c(1, 2, 3, 1, 2, 3, 1, 2, 3), y=c(3, 3, 3, 2, 2, 2, 1, 1, 1), label=1:9) +
    scale_color_manual(values = use_cols,guide="none") +
    scale_fill_manual(values = use_cols,guide="none") +
    theme_classic() +
    xlab("Prop Clear Bins (Bench/Clinical)") +
    ylab("Prop Clear Bins (DM/Comp)")
    
p1 + p2 + plot_layout(design="1122")

```

# Figure 5c (table)

```{r}

select(highlight_points, letter, topic, statement, 
                                 `AvgScore_Bench/Clinical`, `AvgScore_DM/Comp`)

```
